Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis.

### 2024-01-04 paper/paper-new.tex
update

Despite the actual modeling proposal by Liddell and Kruschke (2018), there is a class of regression models taking into account the ordinal nature of the response variable without metric assumptions. We can name this general class of models as \emph{ordinal regression}. The actual statistical nomenclature can be confusing mainly because there are several types of models with different assumptions and structures (Gerhard Tutz, 2022). Gerhard Tutz (2022) and Bürkner and Vuorre (2019) provide a clear and updated taxonomy of ordinal regression models. We can identify three main classes: \emph{cumulative models} {[}CM; McCullagh (1980); Agresti (2010){]}, \emph{sequential models} (Gerhard Tutz, 1990) and \emph{adjacent category models}. The cumulative is the mostly used model assuming the existence of a latent variable that categorized using a set of thresholds produces the observed ordinal variable.

The \emph{sequential model} as suggested by the name is appropriate when modelling sequential processes. Assuming to have five response options, the sequential model assume that responding ``3'' assume a sequential process where steps ``1'' and ``2'' are already reached. A clear example is proposed by Bürkner and Vuorre (2019) where the marriage duration in years is predicted as a function of some explanatory variables. For each level of the response variable there is a latent distribution where the step between a marriage year \(k = 1\) and the next years \(k > 1\) is modeled by the sequential model. When comparing \(k\) with \(k > 1\), everything lower than \(k\) is assumed to be already reached (Gerhard Tutz \& Berger, 2020). The adjacent category model compare the category \(k\) with \(k + 1\) still assuming a latent distribution for each \(k\). As suggested by Gerhard Tutz (2022) the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories. Bürkner and Vuorre (2019) suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical distinction as for the cumulative vs sequential model.

In the current paper we put the focus on the CM for several reasons. The first reason is that the latent formulation of the model is particularly convenient both for parameter interpretation and data simulation. The second reason is that several psychological variables can be formalized as a latent continuous variable observed as an ordinal item. Furthermore, CM are also used to model data under a signal detection theory framework (e.g., DeCarlo, 2010). The Figure \ref{fig:fig-ordinal-models} depict the overall structure of the \emph{cumulative} model.

{\centering \includegraphics[width=1\linewidth]{img/fig-ordinal-models}

\caption{Cumulative model. Adapted from Bürkner and Vuorre (2019).}\label{fig:fig-ordinal-models}
\end{figure}

@@

The \(\mathbf{X} \boldsymbol{\beta}\) is the linear predictor \(\eta\) that is the cumulative probability \(P(Y \leq k)\) transformed using the link function \(g()\). To obtain the probability of a single outcome \(P(Y = k)\) we can compute the difference between cumulative probabilities as shown in Equation \eqref{eq:eq:prob-cum-model2}.

@@

The crucial part is \(\epsilon_i\) that is the random component of the model coming from a certain probability distribution. For a \emph{probit} model, errors are sampled from a standard Gaussian distribution while for a \emph{logit} model from a standard logistic distribution. Following the notation by Gerhard Tutz (2022), the observed ordinal value \(Y_i = k\) comes from \(Y^\star_i\) belonging to the interval defined by the thresholds \(Y_i = k \iff \alpha_{k - 1} < Y^\star_i < \alpha_{k}\) where \(- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty\).

In the basic version of the model, the thresholds \(\alpha_k\) are considered as fixed and being part of the measurement procedure (Liddell \& Kruschke, 2018) and do not vary as a function of the predictors. In a more sophisticated version of the model called location-shift (Gerhard Tutz, 2022), both the location \(\mu\) and the thresholds \(\alpha_k\) can vary as a function of the predictors. Gerhard Tutz (2022) described also another version of the model called location-scale (Cox, 1995; Rigby \& Stasinopoulos, 2005; Gerhard Tutz, 2022) where the location \(\mu\) and the scale \(\sigma^2\) of the distribution can vary as a function of the predictors.

@@

\begin{verbatim}
##   group mean median   sd                  Y1                  Y2
## a     a 2.46      2 1.20 **15** (*p = 0.30*) **11** (*p = 0.22*)
## b     b 3.26      4 0.94  **2** (*p = 0.04*) **11** (*p = 0.22*)
##                    Y3                  Y4
## a **10** (*p = 0.20*) **14** (*p = 0.28*)
## b  **9** (*p = 0.18*) **28** (*p = 0.56*)
\end{verbatim}

@@

@@

\begin{verbatim}
## formula: y ~ x
## data:    dat
## 
##  link  threshold nobs logLik  AIC    niter max.grad cond.H 
##  logit flexible  100  -125.31 258.63 5(0)  3.87e-12 2.1e+01
## 
## Coefficients:
##    Estimate Std. Error z value Pr(>|z|)    
## xb   1.3088     0.3828   3.419 0.000629 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Threshold coefficients:
##     Estimate Std. Error z value
## 1|2  -1.0373     0.3074  -3.374
## 2|3   0.1947     0.2791   0.698
## 3|4   1.0282     0.2947   3.489
\end{verbatim}

The two main sections of the model summary are the \emph{Coefficients} section reporting the regression coefficients \(\beta\) and the \emph{Threshold} section reporting the \(\alpha\) estimation. Given that \(k = 4\) we have \(k - 1 = 3\) thresholds and a single \(\beta\) associated with the \(x\) effect. As in standard regression models, when \(x\) is a categorical predictor with \(j\) level, we will estimate \(j - 1\) regression coefficients where the interpretation depends on the contrast coding {[}for an overview about contrasts coding schemes see Schad, Vasishth, Hohenstein, and Kliegl (2020){]}\footnote{In R the default is the dummy coding where a factor of \(j\) levels is converted into \(j - 1\) dummy variables. By default, the first level of the factor is taken as the reference level and the \(j - 1\) coefficients represent the comparison between the other levels and the reference.}

@@

\caption{(ref:fig-odds-example)}\label{fig:fig-odds-example}
\end{figure}

@@

\CommentTok{\# probability of Y = 1 for the group a}
\NormalTok{dat}\SpecialCharTok{$}\NormalTok{yn }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y) }\CommentTok{\# better as number here}
\NormalTok{(p1a }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
\DocumentationTok{\#\# [1] 0.3}

\CommentTok{\# thus P(Y \textgreater{} 1) = 1 {-} p1a}
\FunctionTok{odds}\NormalTok{(p1a)}
\DocumentationTok{\#\# [1] 0.4285714}

\CommentTok{\# odds ratio Y = 1 a vs b}
\NormalTok{(p1b }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
\DocumentationTok{\#\# [1] 0.04}
\FunctionTok{odds}\NormalTok{(p1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(p1b)}
\DocumentationTok{\#\# [1] 10.28571}
\end{Highlighting}
\end{Shaded}

@@

For example the probability of responding \(Y = 1\) in the group ``a'' is 0.30 corresponding to an odds of 0.43. When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y = 1\) we obtain that the group ``a'' has 10.29 times the odds of responding \(Y = 1\) compared to the group ``b''.

For an ordinal variable we have \(k\) levels and we can calculate \(k - 1\) cumulative probabilities (excluding the last associated with \(p = 1\)). Thus we have \(k - 1\) odds ratio or what is called a cumulative odds ratio. Basically we repeat the previous steps but on the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\). The three odds ratios are very similar correspond to the model estimate \(e^\beta_1\), in R \texttt{exp(fit\$beta)} 3.70. In fact, the logit model essentially estimate the cumulative odds ratio.

@@

\CommentTok{\# calculating the cumulative probability for k {-} 1 variables. }
\CommentTok{\# Taking the average of a series of 0{-}1 is the same as computing the proportion of 1s.}
\NormalTok{(group\_a }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean))}
\DocumentationTok{\#\# y1vs234 y12vs34 y123vs4 }
\DocumentationTok{\#\#    0.30    0.52    0.72}
\NormalTok{(group\_b }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean))}
\DocumentationTok{\#\# y1vs234 y12vs34 y123vs4 }
\DocumentationTok{\#\#    0.04    0.26    0.44}

\CommentTok{\# calculating k {-} 1 odds ratios on the cumulative probabilities as a/b}
\FunctionTok{odds}\NormalTok{(group\_a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(group\_b)}
\DocumentationTok{\#\#   y1vs234   y12vs34   y123vs4 }
\DocumentationTok{\#\# 10.285714  3.083333  3.272727}
\end{Highlighting}
\end{Shaded}

@@

From the previous example, we noticed that the \(k - 1\) odds ratios are very similar (the similarity increase if \(n\) increase). This is not a coincidence but an implicit (until now) assumption made by the CM called \emph{proportional odds} (POA). Following again the taxonomy by Gerhard Tutz (2022), each of the presented ordinal regression model has a basic version making the (POA). There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds}) and partially (\emph{partial proportional odds}). The proportional odds assumption is a crucial point when interpreting model parameters. Can be formalized as in Equation ().

@@

Given the relatively strong assumption, there are several methods for testing if data are supporting the POA (see Liu, He, Tu, \& Tang, 2023 for an overview). Gerhard Tutz and Berger (2020) suggested a trade-off between assuming/relaxing the POA in terms of implementing \emph{location-shift} or \emph{location-scale} models. Basically these methods should guarantee more flexibility in modelling the observed probabilities reducing the number of parameters. However, these methods and models are outside the scope of the tutorial. In the next section, we provide more details only about the \emph{location-scale} model in terms of parameters interpretation and data simulation.

@@

\CommentTok{\# extracting the predicted probabilities for the two groups}
\NormalTok{pr }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(fit, }\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{unique}\NormalTok{(x)))}\SpecialCharTok{$}\NormalTok{fit}

@@

\CommentTok{\# odds ratio y \textless{}= 1 (a vs b) VS y \textless{}= 2 (a vs b)}
\FunctionTok{odds}\NormalTok{(y1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(y1b)}
\DocumentationTok{\#\# [1] 3.701591}
\FunctionTok{odds}\NormalTok{(y12a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(y12b)}
\DocumentationTok{\#\# [1] 3.701591}
\end{Highlighting}
\end{Shaded}

@@

The proportional odds assumption is relevant only for the \emph{logit} model because of the link function that put \(\beta\) as log odds. However the concept of the proportional odds that is more generally called parallel slopes assumption is assumed also for the probit model. Equation \eqref{eq:probit-prop-odds} depict the parallel slopes assumption for a probit model. Basically the difference in \(z\) for a unit increase in \(x\) (i.e., the slope) is the same regardless the \(Y\) level. This can be directly demonstrated in the same way as for the logit model. We use the same dataset presented in Table \ref{tab:tab-prop-ord-ex} but fitting a probit model. The difference between the \(z\) scores comparing the two groups calculated on the cumulative probability is the same regardless the \(k - 1\) level.

\[
z_{x_1 - x_0} = \Phi(P \leq k|x_0) - \Phi(P \leq k|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
\]

@@

\CommentTok{\# z score difference y \textless{}= 1 (a vs b) VS y \textless{}= 2 (a vs b)}
\FunctionTok{qnorm}\NormalTok{(y1a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y1b)}
\DocumentationTok{\#\# [1] 0.8261077}
\FunctionTok{qnorm}\NormalTok{(y12a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y12b)}
\DocumentationTok{\#\# [1] 0.8261077}
\end{Highlighting}
\end{Shaded}

@@

The main difference regards the parameters interpretation. In the logit model the \(\beta\) are the log odds ratio. For categorical variables they represents the increase in the log odds of moving from one level to the other while for numerical variables is the increase in the log odds for a unit increase in \(x\). For the probit model, the \(\beta\) is the increase in terms of \(z\) scores for a unit increase in \(x\). This is very convenient especially for categorical variables because parameters can be intepreted as a Cohen's \(d\) like measure. Thinking about the latent distributions, the \(\beta\) is the shift in the latent mean comparing two or more groups or the slope of latent scores as a function of a numeric \(x\). The intepretation in terms of shifting the latent mean holds also for the logistic model. However, the standard deviation of the standard logistic regression is \(\frac{\pi}{\sqrt{3}} \approx 1.81\)\footnote{Actually the variance of the logistic distribution is \(\frac{s^2\pi^2}{3}\) and the standard deviation \(\frac{s\pi}{\sqrt{3}}\) where \(s\) is the scale of the distribution. For the standard logistic distribution \(s = 1\) (as for the standard normal distribution). For the standard normal distribution \(s^2 = \sigma^2 = 1\)}. The \(\beta\) for the logistic distribution can be interpreted as the location shift of the latent logistic distribution by \(\frac{\pi}{\sqrt{3}}\) standard deviations.

Another advantage of the probit model is the possibility to directly map parameters from signal detection theory (Green, Swets, \& Others, 1966; Stanislaw \& Todorov, 1999) into the model coefficients. In practical terms, the thresholds \(\alpha\) are the criterion cut-offs and the \(\beta_1\) is the d\('\) parameter (DeCarlo, 1998; Knoblauch \& Maloney, 2012).

qualcosa in più qui

@@

There are mainly two ways to simulate data. The first method concerns simulating starting from the latent formulation of the ordinal model. Basically we can simulate the underlying latent distribution and then fixing the thresholds converting the latent continuous variable into the observed ordinal variable.

@@

The first method to simulate ordinal data as a function of predictors is by calculating the true probabilities \(g^{-1}(\eta)\) as a function of predictors and then sample from a multinomial distribution\footnote{In R we are using the \texttt{sample()} function that simulate values from a \emph{categorical} distribution. The \emph{categorical} distribution is considered a special case of the multinomial distribution}. This is similar to the general method to simulate data for a generalized linear model where the linear predictors \(\eta\) are calculated and data are generated from the corresponding probability distribution of the random component. The following code box simulate a \(n\) binary trials with a continuous predictor \(x\).

@@

{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-6-1}

\caption{ }\label{fig:unnamed-chunk-6}
\end{figure}

Essentially the \texttt{lp} (\(\eta_i\)) vector contains true probability of success (\(p_i|x_i = g^{-1}(\eta_i)\)) for the \(x_i\) level of the predictor. Then to introduce the random part of the model, we can use \(p_i\) to generate binary values from a Binomial (in this specific case a Bernoulli) distribution. This create \texttt{y} that is the vector of 0-1 values generated using the true probabilities of success. The supplementary materials contains a more extensive example for simulating data for generalized linear models.

>>>>

Despite the actual modeling proposal by Liddell and Kruschke (2018), there is a class of regression models taking into account the ordinal nature of the response variable without metric assumptions. We can name this general class of models as \emph{ordinal regression}. The actual statistical nomenclature can be confusing mainly because there are several types of models with different assumptions and structures (Tutz, 2022). Tutz (2022) and Bürkner and Vuorre (2019) provide a clear and updated taxonomy of ordinal regression models. We can identify three main classes: \emph{cumulative models} (Agresti, 2010; CM, McCullagh, 1980), \emph{sequential models} (Tutz, 1990) and \emph{adjacent category models}. The cumulative is the mostly used model assuming the existence of a latent variable that categorized using a set of thresholds produces the observed ordinal variable.

The \emph{sequential model} as suggested by the name is appropriate when modelling sequential processes. Assuming to have five response options, the sequential model assume that responding ``3'' assume a sequential process where steps ``1'' and ``2'' are already reached. A clear example is proposed by Bürkner and Vuorre (2019) where the marriage duration in years is predicted as a function of some explanatory variables. For each level of the response variable there is a latent distribution where the step between a marriage year \(k = 1\) and the next years \(k > 1\) is modeled by the sequential model. When comparing \(k\) with \(k > 1\), everything lower than \(k\) is assumed to be already reached (Tutz \& Berger, 2020). The adjacent category model compare the category \(k\) with \(k + 1\) still assuming a latent distribution for each \(k\). As suggested by Tutz (2022) the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories. Bürkner and Vuorre (2019) suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical distinction as for the cumulative vs sequential model.

In the current paper we put the focus on the CM for several reasons. The first reason is that the latent formulation of the model is particularly convenient both for parameter interpretation and data simulation. The second reason is that several psychological variables can be formalized as a latent continuous variable observed as an ordinal item. Furthermore, CM are also used to model data under a signal detection theory framework (e.g., DeCarlo, 2010). The Figure \ref{fig:fig-explain-cumulative} depict the overall structure of the \emph{cumulative} model.

@@

The \(\mathbf{X} \boldsymbol{\beta}\) is the linear predictor \(\eta\) that is the cumulative probability \(P(Y \leq k)\) transformed using the link function \(g()\). To obtain the probability of a single outcome \(P(Y = k)\) we can compute the difference between cumulative probabilities as shown in Equation \eqref{eq:prob-cum-model2}.

@@

The crucial part is \(\epsilon_i\) that is the random component of the model coming from a certain probability distribution. For a \emph{probit} model, errors are sampled from a standard Gaussian distribution while for a \emph{logit} model from a standard logistic distribution. Following the notation by Tutz (2022), the observed ordinal value \(Y_i = k\) comes from \(Y^\star_i\) belonging to the interval defined by the thresholds \(Y_i = k \iff \alpha_{k - 1} < Y^\star_i < \alpha_{k}\) where \(- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty\).

In the basic version of the model, the thresholds \(\alpha_k\) are considered as fixed and being part of the measurement procedure (Liddell \& Kruschke, 2018) and do not vary as a function of the predictors. In a more sophisticated version of the model called location-shift (Tutz, 2022), both the location \(\mu\) and the thresholds \(\alpha_k\) can vary as a function of the predictors. Tutz (2022) described also another version of the model called location-scale (Cox, 1995; Rigby \& Stasinopoulos, 2005; Tutz, 2022) where the location \(\mu\) and the scale \(\sigma^2\) of the distribution can vary as a function of the predictors.

@@

\caption{\label{tab:tab-dataset-example}caption here}
\centering
\begin{tabular}[t]{llllllll}
\toprule
group & mean & median & sd & Y1 & Y2 & Y3 & Y4\\
\midrule
a & 2.42 & 2 & 1.13 & 14 (p = 0.28) & 12 (p = 0.24) & 13 (p = 0.26) & 11 (p = 0.22)\\
b & 3 & 3 & 1.16 & 9 (p = 0.18) & 6 (p = 0.12) & 11 (p = 0.22) & 24 (p = 0.48)\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
\FunctionTok{summary}\NormalTok{(fit)}
\CommentTok{\#\textgreater{} formula: y \textasciitilde{} x}
\CommentTok{\#\textgreater{} data:    dat}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{}  link  threshold nobs logLik  AIC    niter max.grad cond.H }
\CommentTok{\#\textgreater{}  logit flexible  100  {-}132.25 272.51 4(0)  1.08e{-}08 2.0e+01}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Coefficients:}
\CommentTok{\#\textgreater{}    Estimate Std. Error z value Pr(\textgreater{}|z|)   }
\CommentTok{\#\textgreater{} xb   0.9587     0.3718   2.579  0.00992 **}
\CommentTok{\#\textgreater{} {-}{-}{-}}
\CommentTok{\#\textgreater{} Signif. codes:  0 \textquotesingle{}***\textquotesingle{} 0.001 \textquotesingle{}**\textquotesingle{} 0.01 \textquotesingle{}*\textquotesingle{} 0.05 \textquotesingle{}.\textquotesingle{} 0.1 \textquotesingle{} \textquotesingle{} 1}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Threshold coefficients:}
\CommentTok{\#\textgreater{}     Estimate Std. Error z value}
\CommentTok{\#\textgreater{} 1|2  {-}0.8103     0.2814  {-}2.879}
\CommentTok{\#\textgreater{} 2|3   0.0672     0.2636   0.255}
\CommentTok{\#\textgreater{} 3|4   1.1158     0.2900   3.848}
\end{Highlighting}
\end{Shaded}

@@

The two main sections of the model summary are the \emph{Coefficients} section reporting the regression coefficients \(\beta\) and the \emph{Threshold} section reporting the \(\alpha\) estimation. Given that \(k = 4\) we have \(k - 1 = 3\) thresholds and a single \(\beta\) associated with the \(x\) effect. As in standard regression models, when \(x\) is a categorical predictor with \(j\) level, we will estimate \(j - 1\) regression coefficients where the interpretation depends on the contrast coding {[}see Schad, Vasishth, Hohenstein, and Kliegl (2020){]}\footnote{In R the default is the dummy coding where a factor of \(j\) levels is converted into \(j - 1\) dummy variables. By default, the first level of the factor is taken as the reference level and the \(j - 1\) coefficients represent the comparison between the other levels and the reference.}

@@

\caption{caption here}\label{fig:fig-odds-example}
\end{figure}

@@

\CommentTok{\# probability of Y = 1 for the group a}
\NormalTok{dat}\SpecialCharTok{$}\NormalTok{yn }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y) }\CommentTok{\# better as number here}
\NormalTok{(p1a }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
\CommentTok{\#\textgreater{} [1] 0.28}

\CommentTok{\# thus P(Y \textgreater{} 1) = 1 {-} p1a}
\FunctionTok{odds}\NormalTok{(p1a)}
\CommentTok{\#\textgreater{} [1] 0.3888889}

\CommentTok{\# odds ratio Y = 1 a vs b}
\NormalTok{(p1b }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
\CommentTok{\#\textgreater{} [1] 0.18}
\FunctionTok{odds}\NormalTok{(p1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(p1b)}
\CommentTok{\#\textgreater{} [1] 1.771605}
\end{Highlighting}
\end{Shaded}

@@

For example the probability of responding \(Y = 1\) in the group ``a'' is 0.28 corresponding to an odds of 0.39. When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y = 1\) we obtain that the group ``a'' has 1.77 times the odds of responding \(Y = 1\) compared to the group ``b''.

For an ordinal variable we have \(k\) levels and we can calculate \(k - 1\) cumulative probabilities (excluding the last associated with \(p = 1\)). Thus we have \(k - 1\) odds ratio or what is called a cumulative odds ratio. Basically we repeat the previous steps but on the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\). The three odds ratios are very similar correspond to the model estimate \(e^\beta_1\), in R \texttt{exp(fit\$beta)} 2.61. In fact, the logit model essentially estimate the cumulative odds ratio.

@@

\CommentTok{\# calculating the cumulative probability for k {-} 1 variables. }
\CommentTok{\# Taking the average of a series of 0{-}1 is the same as computing the proportion of 1s.}
\NormalTok{(group\_a }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean))}
\CommentTok{\#\textgreater{} y1vs234 y12vs34 y123vs4 }
\CommentTok{\#\textgreater{}    0.28    0.52    0.78}
\NormalTok{(group\_b }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean))}
\CommentTok{\#\textgreater{} y1vs234 y12vs34 y123vs4 }
\CommentTok{\#\textgreater{}    0.18    0.30    0.52}

\CommentTok{\# calculating k {-} 1 odds ratios on the cumulative probabilities as a/b}
\FunctionTok{odds}\NormalTok{(group\_a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(group\_b)}
\CommentTok{\#\textgreater{}  y1vs234  y12vs34  y123vs4 }
\CommentTok{\#\textgreater{} 1.771605 2.527778 3.272727}
\end{Highlighting}
\end{Shaded}

@@

From the previous example, we noticed that the \(k - 1\) odds ratios are very similar (the similarity increase if \(n\) increase). This is not a coincidence but an implicit (until now) assumption made by the CM called \emph{proportional odds} (POA). Following again the taxonomy by Tutz (2022), each of the presented ordinal regression model has a basic version making the (POA). There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds}) and partially (\emph{partial proportional odds}). The proportional odds assumption is a crucial point when interpreting model parameters. Can be formalized as in Equation ().

@@

Given the relatively strong assumption, there are several methods for testing if data are supporting the POA (see Liu, He, Tu, \& Tang, 2023 for an overview). Tutz and Berger (2020) suggested a trade-off between assuming/relaxing the POA in terms of implementing \emph{location-shift} or \emph{location-scale} models. Basically these methods should guarantee more flexibility in modelling the observed probabilities reducing the number of parameters. However, these methods and models are outside the scope of the tutorial. In the next section, we provide more details only about the \emph{location-scale} model in terms of parameters interpretation and data simulation.

@@

\CommentTok{\# extracting the predicted probabilities for the two groups}
\NormalTok{pr }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(fit, }\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{unique}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{x)))}\SpecialCharTok{$}\NormalTok{fit}

@@

\CommentTok{\# odds ratio y \textless{}= 1 (a vs b) VS y \textless{}= 2 (a vs b)}
\FunctionTok{odds}\NormalTok{(y1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(y1b)}
\CommentTok{\#\textgreater{} [1] 2.608304}
\FunctionTok{odds}\NormalTok{(y12a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(y12b)}
\CommentTok{\#\textgreater{} [1] 2.608304}
\end{Highlighting}
\end{Shaded}

@@

The proportional odds assumption is relevant only for the \emph{logit} model because of the link function that put \(\beta\) as log odds. However the concept of the proportional odds that is more generally called parallel slopes assumption is assumed also for the probit model. Equation \eqref{eq:probit-prop-odds} depict the parallel slopes assumption for a probit model. Basically the difference in \(z\) for a unit increase in \(x\) (i.e., the slope) is the same regardless the \(Y\) level. This can be directly demonstrated in the same way as for the logit model. We use the same dataset of the previous simulation. The difference between the \(z\) scores comparing the two groups calculated on the cumulative probability is the same regardless the \(k - 1\) level.

\begin{equation}
z_{x_1 - x_0} = \Phi(P \leq k|x_0) - \Phi(P \leq k|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
\label{eq:probit-prop-odds}
\end{equation}

@@

\CommentTok{\# z score difference y \textless{}= 1 (a vs b) VS y \textless{}= 2 (a vs b)}
\FunctionTok{qnorm}\NormalTok{(y1a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y1b)}
\CommentTok{\#\textgreater{} [1] 0.5644823}
\FunctionTok{qnorm}\NormalTok{(y12a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y12b)}
\CommentTok{\#\textgreater{} [1] 0.5644823}
\end{Highlighting}
\end{Shaded}

@@

The main difference regards the parameters interpretation. In the logit model the \(\beta\) are the log odds ratio. For categorical variables they represents the increase in the log odds of moving from one level to the other while for numerical variables is the increase in the log odds for a unit increase in \(x\).

For the \emph{probit} model, the \(\beta\) is the increase in terms of \(z\) scores for a unit increase in \(x\). This is very convenient especially for categorical variables because parameters can be interpreted as a Cohen's \(d\) like measure. Thinking about the latent distributions, the \(\beta\) is the shift in the latent mean comparing two or more groups or the slope of latent scores as a function of a numeric \(x\). More formally the shift in the latent distribution is \(\frac{\beta}{\sigma}\) (for the \emph{probit} model \(\sigma = 1\)). The interpretation in terms of shifting the latent mean holds also for the logistic model. However, the standard deviation of the standard logistic regression is \(\frac{\pi}{\sqrt{3}} \approx 1.81\)\footnote{Actually the variance of the logistic distribution is \(\frac{s^2\pi^2}{3}\) and the standard deviation \(\frac{s\pi}{\sqrt{3}}\) where \(s\) is the scale of the distribution. For the standard logistic distribution \(s = 1\) (as for the standard normal distribution).}. The \(\beta\) for the logistic distribution can be interpreted as the location shift of the latent logistic distribution by \(\beta/(\frac{\pi}{\sqrt{3}})\) standard deviations.

@@

In this tutorial we present two method for simulating ordinal data. The first method concerns calculating the probabilities of each \(k\) of \(Y\) as a function of predictors and then sampling from a multinomial distribution. The second method simulate the latent variable \(Y^\star\) as a function of predictors.

@@

For the first method we need to calculate \(g^{-1}(\eta)\) as a function of predictors and then sample from a multinomial distribution\footnote{In R we are using the \texttt{sample()} function that simulate values from a \emph{categorical} distribution. The \emph{categorical} distribution is considered a special case of the multinomial distribution}. This is similar to the general method to simulate data for a generalized linear model where the linear predictor \(\eta\) are calculated and data are generated from the corresponding probability distribution of the random component.

As a simple example, let's simulate two groups or condition with \(n = 100\) participants responding to an item with \(k = 4\) ordered options. We simulate that the second group has more probability of responding higher categories of \(Y\). As explained in the previous sections, we can summarise the effect size of a CM (assuming POA) using a single \(\beta = \log(OR)\). We can assume that the odds ratio is \(OR = 2\). For simplicity, the probabilities of the first group \(x = 0\) are uniform thus \(P(Y = 1|x_0) = ... P(Y = k|x_0) = 1/k\). The following code summarise the first steps of the simulation. Basically we define the simulation parameters, we calculate the \(k - 1\) thresholds \(\alpha\) and we apply Equations \eqref{eq:prob-cum-model1} and \eqref{eq:prob-cum-model2}. In this way we calculated the true probability of each \(Y\) level for the two groups.

An important step is converting from \(\alpha\) to probability and the opposite. This steps are implemented in the \texttt{alpha\_to\_prob()} and \texttt{prob\_to\_alpha()} functions. Basically given the input and the link function, these functions return the thresholds associated with \(k\) probabilities or the \(k - 1\) thresholds.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{k }\OtherTok{\textless{}{-}} \DecValTok{5}
\NormalTok{(p }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{k, k)) }\CommentTok{\# uniform}
\CommentTok{\#\textgreater{} [1] 0.2 0.2 0.2 0.2 0.2}
\FunctionTok{names}\NormalTok{(p) }\OtherTok{\textless{}{-}} \FunctionTok{paste0}\NormalTok{(}\StringTok{"y"}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\NormalTok{k)}
